## TODO: 

- If data not successfully generated, should remove the empty .yml file to avoid error in `dataSearch()` -- done!
- reorganize recipes. For reference genome download/indexing, use ensembl, 1000 genomes, not gencode. (example 1, example 5)
- `STAR_index.R` rewrite use `recipeMake` function. SPECIAL CASE: inputparam type=list(), position, output type Directory, glob $(inputs.genomeDir). 
- add meta for gencode transcript recipe. Demo use `gencode_annotation.R`, and `ensembl_liftover.R`. -- done!
- add any new recipe? 
- organize demo for `getData`. 
- Special cases: 
  - take existing `script.sh`.
  - recipeMake arguments types. 

```{R}
devtools::load_all()
library(Rcwl)
```

# For developers: update/search/make recipe, generate data.  

## recipeUpdate take private GitHub repo

```{r recipeUpdate_private_repo}
recipeUpdate(cachePath = "ReUseDataRecipe-private", repo="Liubuntu/ReUseDataRecipe_private")  ## works!
recipeSearch("gencode", cachePath="ReUseDataRecipe-private")  ## works!
```

## Example 1: reference genome GRCh38, Homo sapiens, MT
```{r recipeMake_ref21} 
script <- "
chr=$1
wget http://ftp.ensembl.org/pub/release-104/fasta/homo_sapiens/dna/Homo_sapiens.GRCh38.dna.chromosome.$chr.fa.gz
gzip -d Homo_sapiens.GRCh38.dna.chromosome.$chr.fa.gz 
samtools faidx Homo_sapiens.GRCh38.dna.chromosome.$chr.fa
bwa index Homo_sapiens.GRCh38.dna.chromosome.$chr.fa
"
rcp_ref <- recipeMake(shscript = script,
                      paramID = c("chr"),
                      paramType = c("string"), ## FIXME: require a data.frame, ID, type, check columns for NA, etc. In R, no long in R, use double? Validate! 2e54.(integer, 53 bits integer precision)
                      outputID = "ref_genome",  ## outputType: File[] in code. 
                      outputGlob = "Homo_sapiens.*",   
                      requireTools = c("samtools=1.3", "bwa")  ## name must exactly match the tool name.
                      )
rcp_ref$chr <- "MT" ## FIXME: validate AEAP!! 
res <- getData(rcp_ref,  ## FIXME: check existence (warning, overwrite=TRUE) automatic directory gen related to that recipe!!!
               outdir = "../SharedData/ensembl_refgenome",
               prefix = "ref_genome_MT",
               notes = c("ensembl", "reference", "genome", "MT", "human"),
               ## version = "v0",
               ## docker = TRUE, ## by default: TRUE. 
               ## cwlArgs = "--tmpdir-prefix tmp/",  ## a fix for my desktop, https://cwl.discourse.group/t/docker-error-invalid-mount-config-for-type-bind/598
               showLog = TRUE)
```

## Example 2: gencode annotation gtf file download and unzip (tested!!!)

```{r recipeMake_gencode}
script <- "
species=$1
version=$2
wget http://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_$species/release_$version/gencode.v$version.annotation.gtf.gz
gzip -d gencode.v$version.annotation.gtf.gz
"
## file.copy("/home/qian/workspace/SharedData/gencode_human_38.sh", "gencode_test.sh")
rcp_gencode_annot <- recipeMake(shscript = script,
                                paramID = c("species", "version"),
                                paramType = c("string", "string"),
                                outputID = "annotation",  ## outputType: by default, File[]
                                outputGlob = "gencode.v*.annotation.gtf"
                                ## , requireTools = c() ## wget, gzip 
                                )

rcp_gencode_annot <- addMeta(cwl = rcp_gencode_annot,
                             label = "gencode annotation",
                             doc = "Download and unzip annotation files from gencode",
                             inputLabels = c("species", "version"),
                             inputDocs = c("'human' or 'mouse'",
                                           paste0("Character string. Case sensitive. ",
                                                  "must match available versions for each species under source URL link. ",
                                                  "e.g., 'M31' (species='mouse'), '42' (species='human') ")),
                             outputLabels = c("annotation"),
                             outputDocs = c("the unzipped annotation file: `gencode.v$version.annotation.gtf`"),
                             extensions = list(`$namespaces` = list(
                                                   url = "<http://ftp.ebi.ac.uk/pub/databases/gencode/>",
                                                   date = Sys.Date(),
                                                   example = paste("rcp <- recipeLoad('gencode_annotation')",
                                                                   "rcp$species <- 'human'",
                                                                   "rcp$version <- '42'",
                                                                   "getData(rcp, outdir = 'data/folder', prefix = 'gencode_annotation_human_42', notes = c('gencode', 'annotation', 'human', '42')", sep="\n")))
                             )

cat(meta2md(rcp_gencode_annot))

## write above code in .R file, and upload to "rworkflow/ReUseDataRecipe" GitHub repo.
rm(rcp_gencode_annot)

recipeUpdate(force=TRUE)
recipeSearch(c("gencode", "annotation"))
rcp <- recipeLoad("gencode_annotation")
inputs(rcp)

rcp$species <- "human"
rcp$version <- "42"
getData(rcp,
        outdir = "../SharedData/gencode_annotation",
        prefix = "gencode_annotation_human_42",
        notes = c("gencode", "human", "42"),
        showLog = TRUE
        ## version = "v0",
        ## docker = TRUE,  ## corresponds to recipeMake(requireTools)
        ## cwlArgs = "--tmpdir-prefix tmp/"
        )

rcp$species <- "mouse"
rcp$version <- "M31"
getData(rcp,
        outdir = "../SharedData/gencode_annotation",
        prefix = "gencode_mouse_m31",
        notes = c("gencode", "mouse", "M31"),
        ## docker = TRUE,
        ## cwlArgs = "--tmpdir-prefix tmp/"
        showlog=TRUE
        )
```

As new data generated, we'll update the data path so new data can be searchable. 
```{r}
dataUpdate(dir = "~/workspace/SharedData")
```

## Example 3: ensembl genome lift-over file download and unzip (tested!!!)

recipeMake: `https://github.com/rworkflow/ReUseDataRecipe/blob/master/ensembl_liftover.R`
```{r recipeMake_liftover_mouse}
script <- "
species=$1
from=$2
to=$3
if [ $species == 'human' ]
then 
  species='homo_sapiens'
elif [ $species == mouse ]
then 
  species='mus_musculus'
fi
wget https://ftp.ensembl.org/pub/assembly_mapping/$species/${from}_to_${to}.chain.gz
gzip -d ${from}_to_${to}.chain.gz"

rcp_liftover <- recipeMake(shscript = script,
                           paramID = c("species", "from", "to"), 
                           paramType = c("string", "string", "string"), 
                           outputID = "liftover",  ## required! 
                           outputGlob = "*.chain")  ## required!!
```

```{r recipeLoadMeta}
recipeUpdate(force=TRUE)
rm(rcp_liftover)
recipeSearch("liftover")
rcp <- recipeLoad("ensembl_liftover")
meta(rcp)
cat(RcwlMeta::meta2md(rcp)) 
```

```{r recipeEvaluation}
inputs(rcp)
rcp$species <- "mouse"
rcp$from <- "NCBIM37"
rcp$to <- "GRCm38"
getData(rcp,
        outdir = "../SharedData/ensembl_liftover",
        ## prefix = "ensembl_liftover_ncbim37_grcm38",
        notes = c("ensembl", "liftover", "NCBIM37", "GRCM38"),
        ## version = "v0",
        ## cwlArgs = "--tmpdir-prefix tmp/",
        ## showLog=TRUE)
        showLog = FALSE)
## generate data: GRCh37 to GRCh38

inputs(rcp)
rcp$species <- "human"
rcp$from <- "GRCh37"
rcp$to <- "GRCh38"
getData(rcp,
        outdir = "../SharedData/ensembl_liftover",
        prefix = "ensembl_liftover_grch37_38",
        notes = c("ensembl", "liftover", "GRCh37", "GRCh38"),  ## "notes" will be pasted with " ", so avoid using blank space for notes
        ## version = "v0",
        ## cwlArgs = "--tmpdir-prefix tmp/",
        showLog=TRUE)

dataUpdate(dir = "~/workspace/SharedData")
dataSearch("liftover")
dataSearch(c("liftover", "grcm38"))
```

## Example 4: gencode transcript -- done!

recipeMake: `https://github.com/rworkflow/ReUseDataRecipe/blob/master/gencode_transcripts.R`

```{r recipeLoad}
recipeUpdate(force=TRUE)
recipeSearch(c("gencode", "trans"))
rcp <- recipeLoad("gencode_transcripts")
meta(rcp)
cat(meta2md(rcp))
```

```{r recipeEvaluation}
## version M31 (uses genome assembly version: GRCm39)  worked!!!
rcp$species <- "mouse"
rcp$version <- "M31" 
res <- getData(rcp, outdir = "../SharedData/gencode_transcriptome",
               prefix = "gencode_transcriptome_mouse_M31",
               notes = c("gencode", "transcripts", "mouse", "M31"),
               showLog = TRUE)  ## do not add "version" if already included in parameters. 

## version M20. (test meta... done!) 
rcp$species <- "mouse"
rcp$version <- "M20" 
res <- getData(rcp, outdir = "../SharedData/gencode_transcriptome",
               prefix = "gencode_transcriptome_mouse_M20",
               notes = c("gencode", "transcripts", "mouse", "M20"),
               showLog = TRUE)  ## do not add "version" if already included in parameters. 

## version 42 (human, uses genome assembly version: GRCh38.p13)
rcp$species <- "human"
rcp$version <- "42" 
res <- getData(rcp, outdir = "../SharedData/gencode_transcriptome",
               prefix = "gencode_transcriptome_human_42",
               notes = c("gencode", "human", "42"),
               showLog = TRUE)

## version 19 (human, uses genome assembly version: GRCh37.p13)
rcp$species <- "human"
rcp$version <- "19"
res <- getData(rcp, outdir = "../SharedData/gencode_transcriptome",
               prefix = "gencode_transcriptome_human_19",
               notes = c("gencode", "human", "19"),
               showLog = TRUE)
```

## Example 5: gencode genome assembly 

NOTE: all are different, should we write separate recipes? 

```{r}
script <- "
species=$1
version=$2
if [[ $version = GRCh38 ]]
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_$species/release_42/$version.primary_assembly.genome.fa.gz
elif [[ $version = GRCh37 ]]
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_$species/release_19/$version.p13.genome.fa.gz
elif [[ $species = mouse ]] && [[ $version = GRCm39]]
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M31/$version.primary_assembly.genome.fa.gz
elif [[ $species = mouse ]] && [[ $version = GRCm38 ]]
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M25/$version.primary_assembly.genome.fa.gz
elif [[ $species = mouse ]] && [[ $version = NCBIM37 ]] 
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M1/$version.genome.fa.gz
"
rcp_gencode_refgenome <- recipeMake(script, 
                                paramID = c("species", "version"),
                                paramType = c("string", "string"),
                                outputID = "transcript",
                                outputGlob = "*.transcripts.fa.gz") ## grep 
## write above into .R script, upload to ReUseDataRecipe github repo

## GRCh38
rcp_gencode_trans$species <- "human"
rcp_gencode_trans$version <- "GRCh38" 
res <- getData(rcp_gencode_trans, outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_human_grch38",
               notes = c("gencode", "human", "GRCh38"),
               showLog = TRUE)  ## do not add "version" if already included in parameters. 
## GRCh37
rcp_gencode_trans$species <- "human"
rcp_gencode_trans$version <- "GRCh37" 
res <- getData(rcp_gencode_trans, outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_human_grch37",
               notes = c("gencode", "human", "GRCh37"),
               showLog = TRUE)  ## do not add "version" if already included in parameters.
```

## Example 6, single shell script without any input parameters. 

Need to change cwlProcess class to take null/InputParamList. 

```{r}
script <- "
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_42/GRCh38.primary_assembly.genome.fa.gz
"
rcp_gencode_grch38 <- recipeMake(script, 
                                 outputID = "genome",
                                 outputGlob = "GRCh38.primary_assembly.genome.fa.gz")

## write above into .R script, upload to ReUseDataRecipe github repo
res <- getData(rcp_gencode_grch38,
               outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_grch38",
               notes = c("gencode", "human", "GRCh38"),
               showLog = TRUE)

script <- "
    wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_19/GRCh37.p13.genome.fa.gz
"
rcp_gencode_grch37 <- recipeMake(script, 
                                 outputID = "genome",
                                 outputGlob = "GRCh37.p13.genome.fa.gz")

## write above into .R script, upload to ReUseDataRecipe github repo
res <- getData(rcp_gencode_grch37,
               outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_grch37",
               notes = c("gencode", "human", "GRCh37"),
               showLog = TRUE)

script <- "
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M31/GRCm39.primary_assembly.genome.fa.gz
"
rcp_gencode_grcm39 <- recipeMake(script, 
                                 outputID = "genome",
                                 outputGlob = "GRCm39.primary_assembly.genome.fa.gz")

res <- getData(rcp_gencode_grcm39,
               outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_grcm39",
               notes = c("gencode", "mouse", "GRCm39"),
               showLog = TRUE)

script <- "
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M25/GRCm38.primary_assembly.genome.fa.gz
"
rcp_gencode_grcm38 <- recipeMake(script, 
                                 outputID = "genome",
                                 outputGlob = "GRCm38.primary_assembly.genome.fa.gz")

res <- getData(rcp_gencode_grcm38,
               outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_grcm38",
               notes = c("gencode", "mouse", "GRCm38"),
               showLog = TRUE)

## todo:
script <- "
  wget https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M1/NCBIM37.genome.fa.gz
"
rcp_gencode_ncbim37 <- recipeMake(script, 
                                 outputID = "genome",
                                 outputGlob = "NCBIM37.primary_assembly.genome.fa.gz")

res <- getData(rcp_gencode_ncbim37,
               outdir = "../SharedData/gencode_genome",
               prefix = "gencode_genome_ncbim37",
               notes = c("gencode", "mouse", "NCBIM37"),
               showLog = TRUE)

```

# For users

step 1: [dataUpdate()], dataSearch(), return ("dataHub"? with) data path. 
step 2: if data not exist, [recipeUpdate()], recipeSearch(), return ("dataHub"? with) filepath in local cache. 

## Using data

update all data records, and search data. 

```{r, eval = FALSE}
datapath <- "~/workspace/SharedData"  ## save data to a shared folder in group
dataUpdate(dir = datapath, outMeta = TRUE) ## search local cache (a shared folder within the group) for all available YAML file (go with each data), return a table for all available data.  
aa <- dataSearch(c("gencode", "human", "38")) ## return a "dataHub" object, showing data path. 
``` 

Internally, `meta_data` was run inside `data_Update`, to generate a
summary table for all available dataset by reading the .yml files.

```{r}
metaData <- meta_data(dir = datapath)
```

## Data not exist

Update data recipes from GitHub (`recipeUpdate`), search data recipes
(`recipeSearch`), generate data (`getData`), and write to shared
folder (needs writing permission).

```{r} 
recipeUpdate()
recipeSearch("gencode")  ## return a "dataHub" with recipe name. (recipes are "rcwlProcess" object)
rcp <- recipeLoad("gencode_gtf")  ## load recipe from local rworkflows/ReUseDataRecipes repo.

rcp
inputs(rcp)

rcp$specie = "human"
rcp$version = "38"

res <- getData(rcp, outdir = datapath,
               prefix = "gencode_human_38", 
               note = c("human", "gencode", "gtf"), ## recommended!
               showLog=TRUE,
               docker = FALSE)
dataSearch(c("gencode", "human", "38"))

gencode_gtf$specie <- "mouse"
gencode_gtf$version <- "M27"
res1 <- getData(gencode_gtf, outdir = datapath,
                prefix = "gencode_mouse_M27",
                note = c("mouse", "gencode", "gtf"),
                showLog = TRUE,
                docker = FALSE)
dataSearch(c("gencode", "mouse", "M27"))

recipeSearch(c("STAR", "index"))
rcp <- recipeLoad("STAR_index")

inputs(rcp)
rcp$ref <- ""
rcp$gtf <- ""
rcp$genomeDir <- "STARindex"
rcp$threads <- 1
rcp$sjdb <- 100

getData(rcp, outdir = datapath,
           prefix = "reference_genome_",
           note = c("reference", "genome", "human"),
           showLog = TRUE)
```

## Recipe not exist

- make a request to administrator. OR 
- write your own script to get data (downloading/curation, etc.),
  referring to the available scripts:
  https://github.com/rworkflow/ReUseDataRecipe


## todo:

1. getData, always print "Note" and "Version"? -- DONE!
2. create a "meta_recipe.csv", searchable by recipeSearch() function.  -- REMOVE!
3. create a "meta_data.csv", searchable by dataSearch() function.  -- DONE! 
   - `meta_data(dir)` can be multiple paths... 
4. "recipeUpdate()": cache existing recipes from GitHub.   -- DONE!
5. "recipeSearch()" to search cached path of recipe. -- DONE!
6. "recipeLoad()", run the recipe as a whole, and generate a cwlProcess object, only load the cwlProcess object into R, assign values, and generate data. -- renamed as "getData"???
7. "dataUpdate" function: cache existing data by providing the datapath, generate a bfc in cachepath (e.g., "ReUseData") for search purpose.   -- DONE! 
8. "dataSearch()" use keywords to search existing data in bfc. -- DONE!
   - exact match? currently case sensitive by `bfcquery()`... 
   
9. create "dataHub" class, extend "cwlHub", only show some info, e.g., rid, fpath, Version, Date, etc.  -- DONE!
10. FIXME update "recipeLoad(cachePath=)", with support of github repo "rworkflows/ReUseDataRecipes". 

## Questions:


## Data aggregate by software
```{r}
dataUpdate("/projects/rpci/shared/references/")
recipeSearch("reference", "ReUseDataRecipe-private")
dat1 <- dataSearch(c("reference","fasta", "hs37d5"))
dat2 <- dataSearch(c("GENCODE"))


## setMethod("[", c("dataHub"), function(x, i) {
##     rids <- x@rid[i]
##     return(x[rids])
## })
## setGeneric("[")

## setMethod("c", c("dataHub"), function(x, ...) {
##     object <- list(x, ...)
##     rids <- unlist(lapply(object, function(x)x@rid))
##     x@rid <- unique(rids)
##     return(x)
## })

## setGeneric("c")

## setMethod("as.list", c("dataHub"), function(x, type = NULL) {
##     tl <- title(x)
##     pth <- dataPath(x)
##     if(!is.null(type) && type == "cwl"){
##         dtype <- unlist(lapply(pth, function(x)file.info(x)$isdir))
##         dtype <- ifelse(dtype, "Directory", "File")
##         dl <- vector("list", length(pth))
##         for(i in 1:length(pth)){
##             dl[[i]] <- list(class = dtype[i],
##                             path = pth[i])
##         }
##     }else{
##         dl <- as.list(dataPath(x))
##     }
##     names(dl) <- title(x)
##     return(dl)
## })
## setGeneric("as.list")

## tag <- function(object){
##     mcols(object)$tag
## }
## "tag<-" <- function(object, value, append = TRUE){
##     ## mcols(dat1)@listData$tag <- value
##     if(append){
##         value <- paste0(tag(object), value)
##     }
##     dm <- bfcmeta(object, "dataMeta")
##     idx <- match(object@rid, dm$rid)
##     dm$tag[idx] <- value
##     bfc <- BiocFileCache(object@cache, ask = FALSE)
##     bfcmeta(bfc, "dataMeta", overwrite = TRUE) <- dm
##     return(object)
## }


## tags <- function(object){
##     mcols(object)[, c("fpath", "tag")]
## }
## #' @param values A data.frame of fpath and tags to update the metadata.
## "tags<-" <- function(object, values){
##     dm <- bfcmeta(object, "dataMeta")
##     rids <- object@rid[match(values$fpath, dataPath(object))]
##     dm$tag[match(rids, dm$rid)] <- values$tag

##     bfc <- BiocFileCache(object@cache, ask = FALSE)
##     bfcmeta(bfc, "dataMeta", overwrite = TRUE) <- dm
##     return(object)
## }

### Example
dat1 <- dataSearch(c("reference","STAR", "GRCh38"))
dat2 <- dataSearch(c("GENCODE"))

## merge for STAR alignment
r1 <- c(dat1[17], dat2[2])

tag(r1, append=F) <- rep("#STAR#GRCh38", 2)
tag(r1)

## use example
r2 <- dataSearch("#STAR")
to_list(r2)
to_list(r2, "cwl")

tags(r2)

## mutect2
script <- "
path=$1
idx=$2
wget https://storage.googleapis.com/$path
if [ ! -z $idx ]; then wget https://storage.googleapis.com/$path.$idx; fi
"
googleBucket <- recipeMake(shscript = script,
                           paramID = c("rpath", "index"),
                           paramType = c("string", "string?"),
                           outputID = "dfile",
                           outputGlob = "$(inputs.rpath.split('/').slice(-1)[0])*")

googleBucket$rpath <- "gatk-best-practices/somatic-b37/Mutect2-exome-panel.vcf"
googleBucket$index <- "idx"
getData(googleBucket,
        outdir = "../SharedData/Mutect2",
        prefix = "Mutect2-exome-panel",
        notes = c("mutect2", "b37", "exome-panel", "pon"), showLog=TRUE)

## googleBucket$rpath <- "gatk-best-practices/somatic-b37/af-only-gnomad.raw.sites.vcf"
## googleBucket$index <- "idx"
## getData(googleBucket,
##         outdir = "../SharedData/Mutect2",
##         prefix = "af-only-gnomad",
##         notes = c("mutect2", "b37", "af-only-gnomad"), showLog=TRUE)

googleBucket$rpath <- "gatk-best-practices/somatic-b37/small_exac_common_3.vcf"
googleBucket$index <- "idx"
getData(googleBucket,
        outdir = "../SharedData/Mutect2",
        prefix = "small_exac_common_3",
        notes = c("mutect2", "b37", "small_exac_common"), showLog=TRUE)

dataUpdate("../SharedData")

r3 <- dataSearch("mutect2")
r3 <- r3[c(1,3)]
tag(r3) <- rep("#mutect2#b37", 2)

r3 <- dataSearch(c("#mutect2", "#b37"))
toList(r3)
toList(r3, "cwl")
```
