---
title: "ReUseData: An R/Bioconductor Data Management Tool for Reusable and Reproducible Biomedical Data Resources"
author: "Qian Liu"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{ReUseData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


`ReUseData` is a _Bioconductor_ package that manages a collection
of commonly used bioinformatics tools and pipeline based on
`Rcwl`. These pre-built and pre-tested tools and pipelines are highly
modularized with easy customization to meet different bioinformatics
data analysis needs.  

`Rcwl` and `ReUseData` together forms a _Bioconductor_ toolchain
for use and development of reproducible bioinformatics pipelines in
Common Workflow Language (CWL). The project also aims to develop a
community-driven platform for open source, open development, and open
review of best-practice CWL bioinformatics pipelines.

Function: 

Data management: 
- getData(), 
- dataUpdate(), dataSearch(), dataHub class, show method, subsetting dataHub[index], combining c(dh, dh), getter funcs: dataNotes(), dataPath(), dataParams(), title(), dataTag()? toList() to use data!!!

Data recipes: 
- recipeHub class, show methods, subsetting recipeHub[index], 
- recipeMake, recipeSearch, recipeLoad
- recipe evaluation: getData

```{r}
recipeUpdate(force=TRUE)
rcps <- recipeSearch(c("gencode"))
rcp1 <- rcps[1]
recipeNames(rcp1)
```

## private GitHub repo? example. 
```{r}
dataSearch()
dd <- dataSearch("reference", "38", "1000g"))
dataNames(dd)
dataParams(dd)
dataNotes(dd)
dataTags(dd)
toList(dd)
toList(dd, format = "json")
toList(dd, format = "yaml")
```

## Data recipe construction and evaluation

recipe construction: 

Convert existing shell scripts for data processing into data recipes,
by specifying input parameters, and output globbing pattern. The the
data recipe is represented in _R_ as an S4 class `cwlProcess`. 

```{r}
shfile <- system.file("extdata", "demo_script.sh", package = "ReUseData")
readLines(shfile)
rcp <- recipeMake(shscript = shfile,
                  paramID = c("species", "version"),
                  paramType = c("string", "string"),
                  outputID = "annotation", 
                  outputGlob = "gencode.v*.annotation.gtf"
                  )
rcp
```

Users can then assign values to the input parameters, and evaluate the
recipe (`getData`) to generate data of interest. 

```{r}
rcp$species <- "human"
rcp$version <- "42"
res <- getData(rcp,
        outdir = tempdir(), 
        prefix = "gencode_annotation_human_42",
        notes = c("gencode", "human", "42"),
        showLog = TRUE)
```

The file path to newly generated dataset can be easily retrieved. 

```{r}
res$output
```

There are some automatically generated files to help track the data
recipe evaluation, including `*.sh` to record the original shell
script, `*.cwl` file as the official workflow script which was
internally submitted for data recipe evaluation, `*.yml` file as part
of CWL workflow evalution, which also record data annotations, and
`*.md5` checksum file to check/verify the integrity of generated data
file.

```{r}
dir(tempdir(), pattern = "gencode_annotation_human_42")
```

The `*.yml` file contains information about recipe input parameters,
the file path to output file, the notes for the dataset, and
auto-added date for data generation time. A later data search using
`dataSearch()` will refer to this file for keywords match.

```{r}
readLines(res$yml)
```

```{r}
script <- '
aa=$1
bb=$2
echo "Print the input: $aa" > $bb.txt
'
rcp <- recipeMake(shscript = script,
                  paramID = c("aa", "bb"),
                  paramType = c("string", "string"),
                  outputID = "echoout",
                  outputGlob = "*.txt")
inputs(rcp)
outputs(rcp)
rcp$aa <- "Hello World!"
rcp$bb <- "outfile"
res <- getData(rcp, outdir = tempdir(),
               prefix = "test",
               notes = c("echo", "txt", "test"),
               showLog = TRUE)
```

Let's take a look at the output file, which is successfully generated
in user-specified directory and grabbed through the `outputGlob`
argument.

```{r}
res$out
readLines(res$out)
```


```{r}
readLines(res$yml)
```

## Data recipe caching and updating

`recipeUpdate()` creates a local cache for data recipes that are saved
in specified GitHub repository (if first time use), syncs and updates
data recipes from the GitHub repo to local caching system, so any
newly added recipes can be readily accessed and loaded directly in
_R_.  

**NOTE:** 

- The `cachePath` argument need to match between `recipeUpdate`,
`recipeLoad` and `recipeSearch` functions.
- use `force=TRUE` when any old recipes that are previously cached are
updated.

```{r}
## First time use
recipeUpdate(cachePath = "ReUseDataRecipe",
             repos = "rworkflow/ReUseDataRecipe",
             force = TRUE)
## afterwards
recipeUpdate()
```

## Data recipe searching and loading

Cached data recipes can be searched using multiple keywords. It
returns a `recipeHub` object with a list of recipes available. 

```{r}
recipeSearch()
recipeSearch("gencode")
recipeSearch(c("STAR", "index"))
```

Recipes can be directly loaded into _R_ using `recipeLoad` function
with the exact recipe name. It's important to check the input
parameters of a data recipe, and assign values before evaluting a
recipe to generate data of interest.

```{r}
rcphub <- recipeSearch("gencode")
recipeNames(rcphub)
rcp1 <- recipeLoad("gencode_annotation")
inputs(rcp)
rcp <- recipeLoad("STAR_index")
inputs(rcp)
```


# Installation
1. Install the package from _Bioconductor_.

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ReUseData")
```
The development version is also available to download from GitHub. 
```{r getDevel, eval=FALSE}
BiocManager::install("rworkflow/ReUseData")
```

2. Load the package into the R session.
```{r Load, message=FALSE}
library(ReUseData)
```
		
# Project resources 

The project website https://rcwl.org/ serves as a central hub for all
related resources. It provides guidance for new users and tutorials
for both users and developers. Specific resources are listed below. 

## The _R_ recipes and cwl scripts

The _R_ scripts to build the CWL tools and pipelines are now residing
in a dedicated [GitHub
repository](https://github.com/rworkflow/RcwlRecipes), which is
intended to be a community effort to collect and contribute
Bioinformatics tools and pipelines using `Rcwl` and CWL.

## Tutorial book

The [tutorial book](https://rcwl.org/RcwlBook/) provides detailed
instructions for developing `Rcwl` tools/pipelines, and also includes
examples of some commonly-used tools and pipelines that covers a wide
range of Bioinformatics data analysis needs.


# `ReUseData` core functions

Here we show the usage of 3 core functions: `cwlUpdate`, `cwlSearch`
and `cwlLoad` for updating, searching, and loading the needed tools
or pipelines in _R_.

## `cwlUpdate` 

The `cwlUpdate` function syncs the current `Rcwl` recipes and returns
a `cwlHub` object which contains the most updated `Rcwl` recipes. The
`mcols()` function returns all related information about each
available tool or pipeline. 

The recipes will be locally cached, so users don't need to call
`cwlUpdate` every time unless they want to use a tool/pipeline that is
newly added to `RcwlPipelines`. Here we are using the recipes from
_Bioconductor_ devel version.
	
```{r}
## For vignette use only. users don't need to do this step.
Sys.setenv(cachePath = tempdir()) 
```

```{r, message=FALSE}
atls <- cwlUpdate(branch = "dev") ## sync the tools/pipelines.
atls
table(mcols(atls)$Type)
```
Currently, we have integrated `r sum(Type(atls)=="tool")` command
line tools and `r sum(Type(atls)=="pipeline")` pipelines.

## `cwlSearch`

We can use (multiple) keywords to search for specific tools/pipelines
of interest, which internally search the `mcols` of "rname", "rpath",
"fpath", "Command" and "Containers". Here we show how to search the
alignment tool `bwa mem`.

```{r}
t1 <- cwlSearch(c("bwa", "mem"))
t1
mcols(t1)
```

## `cwlLoad`

The last core function `cwlLoad` loads the `Rcwl` tool/pipeline into
the _R_ working environment. The code below loads the tool with a
user-defined name `bwa` to do the read alignment. 

```{r}
bwa <- cwlLoad(title(t1)[1])  ## "tl_bwa"
bwa <- cwlLoad(mcols(t1)$fpath[1]) ## equivalent to the above. 
bwa
```

Now the _R_ tool of `bwa` is ready to use. 

# Customize a tool or pipeline

To fit users' specific needs，the existing tool or pipline can be
easily customized. Here we use the `rnaseq_Sf` pipeline to demonstrate
how to access and change the arguments of a specific tool inside a
pipeline. This pipeline covers RNA-seq reads quality summary by
`fastQC`, alignment by `STAR`, quantification by `featureCounts` and
quality control by `RSeQC`.

```{r, warning=FALSE}
rnaseq_Sf <- cwlLoad("pl_rnaseq_Sf")
plotCWL(rnaseq_Sf)
```

There are many default arguments defined for the tool of `STAR` inside
the pipeline. Users might want to change some of them. For example, we
can change the value for `--outFilterMismatchNmax` argument from 2 to
5 for longer reads.

```{r}
arguments(rnaseq_Sf, "STAR")[5:6]
arguments(rnaseq_Sf, "STAR")[[6]] <- 5
arguments(rnaseq_Sf, "STAR")[5:6]
```

We can also change the docker image for a specific tool (e.g., to a
specific version). First, we search for all available docker images
for `STAR` in biocontainers repository. The Source server could be
[quay](https://quay.io/) or [dockerhub](https://hub.docker.com).

```{r}
searchContainer("STAR", repo = "biocontainers", source = "quay")
```

Then, we can change the `STAR` version into 2.7.8a (tag name: 2.7.8a--0). 

```{r}
requirements(rnaseq_Sf, "STAR")[[1]]
requirements(rnaseq_Sf, "STAR")[[1]] <- requireDocker(
    docker = "quay.io/biocontainers/star:2.7.8a--0")
requirements(rnaseq_Sf, "STAR")[[1]]
```

# Run a tool or pipeline

Once the tool or pipeline is ready, we only need to assign values for
each of the input parameters, and then submit using one of the
functions: `runCWL`, `runCWLBatch` and `cwlShiny`. More detailed Usage
and examples can be refer to the `Rcwl`
[vignette](https://bioconductor.org/packages/devel/bioc/vignettes/Rcwl/inst/doc/Rcwl.html).

To successfully run the tool or pipeline, users either need to have
all required command line tools pre-installed locally, or using the
docker/singularity runtime by specifying `docker = TRUE` or `docker =
"singularity"` argument inside `runCWL` or `runCWLBatch`
function. Since the _Bioconductor_ building machine doesn't have all the
tools installed, nor does it support the docker runtime, here we use some
pseudo-code to demonstrate the tool/pipeline execution.

```{r, eval=FALSE}
inputs(rnaseq_Sf)
rnaseq_Sf$in_seqfiles <- list("sample_R1.fq.gz",
                              "sample_R2.fq.gz")
rnaseq_Sf$in_prefix <- "sample"
rnaseq_Sf$in_genomeDir <- "genome_STAR_index_Dir"
rnaseq_Sf$in_GTFfile <- "GENCODE_version.gtf"

runCWL(rnaseq_Sf, outdir = "output/sample", docker = TRUE)
```

Users can also submit parallel jobs to HPC for multiple samples using
`runCWLBatch` function. Different cluster job managers, such as
"multicore", "sge" and "slurm", are supported using the
`BiocParallel::BatchtoolsParam`.

```{r, eval=FALSE}
library(BioParallel)
bpparam <- BatchtoolsParam(workers = 2, cluster = "sge",
                           template = batchtoolsTemplate("sge"))

inputList <- list(in_seqfiles = list(sample1 = list("sample1_R1.fq.gz",
                                                    "sample1_R2.fq.gz"),
                                     sample2 = list("sample2_R1.fq.gz",
                                                    "sample2_R2.fq.gz")),
                  in_prefix = list(sample1 = "sample1",
                                   sample2 = "sample2"))

paramList <- list(in_genomeDir = "genome_STAR_index_Dir",
                  in_GTFfile = "GENCODE_version.gtf",
                  in_runThreadN = 16)

runCWLBatch(rnaseq_Sf, outdir = "output",
            inputList, paramList,
            BPPARAM = bpparam)
```

# SessionInfo
```{r}
sessionInfo()
```

